A Bidirected String Graph Model for Genome Assembly

### 2012-12-12 presentation/paper.tex
small example

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
%\usepackage{amsthm}

@@

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\begin{tikzpicture}=\ReadArrowType]
			\draw[->,style=thick] (0,0) -- (2,0);
			\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
		\end{tikzpicture}
		%\vspace{3mm}
		%or (symmetrically)
		%\vspace{3mm}

		%\begin{tikzpicture}=\ReadArrowType]
			%\draw[->,style=thick] (2,0) -- (0,0);
			%\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
		%\end{tikzpicture}
		\caption{``Normal'' overlap}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\begin{tikzpicture}=\ReadArrowType]
			\draw[->,style=thick] (0,0) -- (2,0);
			\draw[->,style=thick] (3,0.3) -- (1,0.3);
		\end{tikzpicture}
		\caption{``Innie'' overlap}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\begin{tikzpicture}=\ReadArrowType]
			\draw[->,style=thick] (2,0) -- (0,0);
			\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
		\end{tikzpicture}
		\caption{``Outie'' overlap}
	\end{subfigure}
	\caption{The three types of overlaps.  The innie and outie overlaps are
	reverse-complement overlaps, while the normal overlap is not.  Note:
	with the normal overlap, there is a symmetric case that is not
	considered a distinct type of overlap here.}
	\label{fig:overlap_types}
\end{figure}

@@

>>>>

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
\usepackage{grffile}
%\usepackage{amsthm}

@@

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.3\textwidth}
		\begin{tikzpicture}[>=\ReadArrowType]
			\draw[->,style=thick] (0,0) -- (2,0);
			\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
		\end{tikzpicture}
		%\vspace{3mm}
		%or (symmetrically)
		%\vspace{3mm}

		%\begin{tikzpicture}=\ReadArrowType]
			%\draw[->,style=thick] (2,0) -- (0,0);
			%\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
		%\end{tikzpicture}
		\caption{``Normal'' overlap}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\begin{tikzpicture}[>=\ReadArrowType]
			\draw[->,style=thick] (0,0) -- (2,0);
			\draw[->,style=thick] (3,0.3) -- (1,0.3);
		\end{tikzpicture}
		\caption{``Innie'' overlap}
	\end{subfigure}
	\begin{subfigure}[b]{0.3\textwidth}
		\begin{tikzpicture}[>=\ReadArrowType]
			\draw[->,style=thick] (2,0) -- (0,0);
			\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
		\end{tikzpicture}
		\caption{``Outie'' overlap}
	\end{subfigure}
	\caption{The three types of overlaps.  The innie and outie overlaps are
	reverse-complement overlaps, while the normal overlap is not.  Note:
	with the normal overlap, there is a symmetric case that is not
	considered a distinct type of overlap here.}
	\label{fig:overlap_types}
\end{figure}

@@

This section walks through the construction and simplification of the bidirected
string graph on a very small example and provides a more detailed look at the
actual data.  The genome is 200 bp long and is given by the following file:

%make -B SAMPLE_SIZE=200 READ_LEN=50 MIN_OVERLAP_LEN=15 out.reduced.mapped.bidigraph.pdf out.reduced.mapped.collapsed.bidigraph.pdf out.bidigraph.pdf COVERAGE=2.5 SEED=9 USE_PIRS=false

\begin{verbatim}
>random_genome
ACTCGAATTCCGAATAGATAGAAGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTC
GGGGAACGTGGCGGGTAGAGCTATCGTGTCACTCTCTGTCAGTGGTCTGTGAAAATTGAT
CGCTACTGTCCCGGTGTAACCAGTCGACCCGCGGCAACTGCGAAGGGCCTCTCGCTCGCA
GTACTTCTCGGAGCTATCCC
\end{verbatim}

The reads are the following 10 reads (50bp each, sampled randomly from either
strand of the above genome):

\begin{verbatim}
>read_1
CAGAACCGGCGCGATATTATCAGCAGACTTCTATCTATTCGGAATTCGAG
>read_2
GCAGTTGCCGCGGGTCGACTGGTTACACCGGGACAGTAGCGATCAATTTT
>read_3
GCGGGTAGAGCTATCGTGTCACTCTCTGTCAGTGGTCTGTGAAAATTGAT
>read_4
AGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTCGGGGAACGTGGC
>read_5
TTCGCAGTTGCCGCGGGTCGACTGGTTACACCGGGACAGTAGCGATCAAT
>read_6
TACACCGGGACAGTAGCGATCAATTTTCACAGACCACTGACAGAGAGTGA
>read_7
ACAGAGAGTGACACGATAGCTCTACCCGCCACGTTCCCCGAGGCGCATCA
>read_8
ACAGACCACTGACAGAGAGTGACACGATAGCTCTACCCGCCACGTTCCCC
>read_9
TAGCGATCAATTTTCACAGACCACTGACAGAGAGTGACACGATAGCTCTA
>read_10
CTATCGTGTCACTCTCTGTCAGTGGTCTGTGAAAATTGATCGCTACTGTC
\end{verbatim}

The location on the genome from which each of these reads came is shown below.
Note that by chance, 36 bp from the right end of the genome are not included in
any read:

\newcounter{readNumber}
\newcounter{readPos}
\newcounter{forwardReads}
\newcounter{rcReads}

\setcounter{forwardReads}{0}
\setcounter{rcReads}{1}

% ForwardRead{pos}{len}
\newcommand{\ForwardRead}[2]{
	\addtocounter{readNumber}{1}
	\addtocounter{forwardReads}{-1}
	\setcounter{readPos}{#1}
	\addtocounter{readPos}{#2}
	\draw[->, style=thick] (#1, \arabic{forwardReads}) node[anchor=east]
									{\arabic{readNumber}.} -- (\arabic{readPos},
									\arabic{forwardReads});
}

% RcRead{pos}{len}
\newcommand{\RcRead}[2]{
	\addtocounter{readNumber}{1}
	\addtocounter{rcReads}{1}
	\setcounter{readPos}{#1}
	\addtocounter{readPos}{#2}
	\draw[<-, style=thick] (#1, \arabic{rcReads}) node[anchor=east]
									{\arabic{readNumber}.} -- (\arabic{readPos},
									\arabic{rcReads});
}

\begin{center}
	\begin{tabular}{|ccc|}
	\hline
	{\bf Read number} & {\bf Position} & {\bf Reverse-complement?} \\ \hline
	1  & 1  & yes \\
	2  & 111  & yes \\
	3  & 70  & no \\
	4  & 22  & no \\
	5  & 114  & yes \\
	6  & 88  & yes \\
	7  & 49  & yes \\
	8  & 60  & yes \\
	9  & 75  & yes \\
	10  & 80  & no \\ \hline
	\end{tabular}

	\fbox{
	\begin{tikzpicture}[>=\ReadArrowType,x=0.56mm,y=3.5mm]
		\draw[<-, style=thick] (0, 1) node[anchor=east] {G.} -- (200, 1);
		\draw[->, style=thick] (0, 0) node[anchor=east] {G.} -- (200, 0);

		\RcRead{1}{50}
		\RcRead{111}{50}
		\ForwardRead{70}{50}
		\ForwardRead{22}{50}
		\RcRead{114}{50}
		\RcRead{88}{50}
		\RcRead{49}{50}
		\RcRead{60}{50}
		\RcRead{75}{50}
		\ForwardRead{80}{50}
	\end{tikzpicture}
	}
\end{center}

\ProgramName{compute-overlaps} computes the following exact overlaps of minimum
length 15 bp among the above reads:

\begin{verbatim}
Overlap { Read 1: [0, 28], Read 4: [0, 28], rc = 1 }
Overlap { Read 2: [0, 46], Read 5: [3, 49], rc = 0 }
Overlap { Read 2: [23, 49], Read 6: [0, 26], rc = 0 }
Overlap { Read 2: [31, 49], Read 10: [31, 49], rc = 1 }
Overlap { Read 3: [18, 49], Read 6: [18, 49], rc = 1 }
Overlap { Read 3: [0, 28], Read 7: [0, 28], rc = 1 }
Overlap { Read 3: [0, 39], Read 8: [0, 39], rc = 1 }
Overlap { Read 3: [5, 49], Read 9: [5, 49], rc = 1 }
Overlap { Read 3: [10, 49], Read 10: [0, 39], rc = 0 }
Overlap { Read 4: [27, 49], Read 7: [27, 49], rc = 1 }
Overlap { Read 5: [26, 49], Read 6: [0, 23], rc = 0 }
Overlap { Read 5: [34, 49], Read 10: [34, 49], rc = 1 }
Overlap { Read 6: [28, 49], Read 8: [0, 21], rc = 0 }
Overlap { Read 6: [13, 49], Read 9: [0, 36], rc = 0 }
Overlap { Read 6: [8, 49], Read 10: [8, 49], rc = 1 }
Overlap { Read 7: [0, 38], Read 8: [11, 49], rc = 0 }
Overlap { Read 7: [0, 23], Read 9: [26, 49], rc = 0 }
Overlap { Read 7: [0, 18], Read 10: [0, 18], rc = 1 }
Overlap { Read 8: [0, 34], Read 9: [15, 49], rc = 0 }
Overlap { Read 8: [0, 29], Read 10: [0, 29], rc = 1 }
Overlap { Read 9: [0, 44], Read 10: [0, 44], rc = 1 }
\end{verbatim}

These overlaps are then used by \ProgramName{build-bidirected-string-graph} to
build the assembly graph.  Note that there are no contained reads in this
example, so there are as many vertices as original reads.  Also, in this image
(and the following images), each edge is labeled with the length of that edge's
sequence, rather than the sequence itself.  Furthermore, although each edge
really contains a separate sequence for each direction, for this particular
example all the reads are the same length and only exact overlaps are
considered, which makes it so the two sequences of each edge are the same
length.

\begin{center}
	\includegraphics[scale=0.7]{example_200bp.bidigraph-crop.pdf}	
\end{center}

The assembly graph then can be simplified by \ProgramName{transitive-reduction}:

\begin{center}
	\includegraphics[scale=0.8]{example_200bp.reduced.mapped.bidigraph-crop.pdf}	
\end{center}

Then, the assembly can be simplified further by
\ProgramName{collapse-unbranched-paths}:

\begin{center}
	\includegraphics[scale=0.8]{example_200bp.reduced.mapped.collapsed.bidigraph-crop.pdf}	
\end{center}

%\begin{figure}
	%\begin{center}
		%\begin{subfigure}{0.99\textwidth}
			%\begin{center}
				%\includegraphics[scale=0.5]{example.bidigraph-crop.pdf}	
				%\caption{Bidirected string graph built from 100 bp reads sampled
				%randomly at 5 $\times$ coverage from a 450 bp genome, using minimum 25
				%bp overlaps.  Note: edges are labeled with their lengths, not DNA
				%sequences.}
			%\end{center}
		%\end{subfigure}
		%\begin{subfigure}{0.99\textwidth}
			%\begin{center}
				%\includegraphics[scale=0.7]{example.reduced.mapped.bidigraph-crop.pdf}	
				%\caption{The bidirected string graph from the previous slide, after
				%transitive reduction.}
			%\end{center}
		%\end{subfigure}
		%\begin{subfigure}{0.99\textwidth}
			%\begin{center}
				%\includegraphics[scale=0.7]{example.reduced.mapped.collapsed.bidigraph-crop.pdf}	
				%\caption{The bidirected string graph from the previous slide, after
				%collapsing unbranched paths.}
			%\end{center}
		%\end{subfigure}
	%\end{center}
%\end{figure}
